---
title: "kknn commands"
author: "David Oury"
date: "11 Apr 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `caret` package makes available the `confusionMatrix` function
([documentation](https://artax.karlin.mff.cuni.cz/r-help/library/caret/html/confusionMatrix.html).)
The `kknn` package makes available the `kknn` function.
The `e1071` package is required by the `caret` package.
```{r message=FALSE}
install.packages("kknn")
install.packages("caret")
install.packages("e1071")
library(kknn)
library(caret)
library(e1071)

options(dplyr.width=Inf) 
```

We use the `iris` dataset.
```{r}
Data <- iris
head(Data, 10)
dim(Data)
```

```{r}
Sample   <- sample(1:150, 50)
Sample
testing  <- Data[ Sample, ]
training <- Data[-Sample, ]
dim(training)
dim(testing)
```

Train the model on the `training` dataset. 
The target variable is `Species`. 
Look at all values for `k` up to 9.
```{r}
suppressWarnings(suppressMessages(library(kknn)))
model <- train.kknn(Species ~ ., 
                    data = training, 
                    kmax = 9)
model
```

Create predictions for the `testing` dataset (based on the model.)
```{r}
prediction <- predict(model, testing[, -5])
prediction
```

Display details on the correctness of this prediction. 
```{r}

confusionMatrix(reference=testing[, 5], data=prediction)


```

```{r}
plot(model)
```

